---
title: "Covid19 Report"
author: "Som. Thapa"
date: "06/12/2023"
output: html_document
---


## Introduction

This is part of final project for DTSA 5301 Data Science as a Field. Objective of the report is develop essential skills for data analysis ie. tidy, transform, analyze and report the conclusion
 
### Goal of the Project

The goal of this project is to analyze how COVID-19 spreads over time over the world and in United States. In addition, I will also explore if there is any relation between COVID-19 deaths and impact of vaccinations


### Import R libraries and set up environment

```{r setup,message=FALSE,warning=FALSE}
# Import libraries and set options
library(tidyverse)
library(lubridate)
options(warn=-1)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = getwd())
```

## Data Import

### Data Source

The report makes use of publicly available data from John Hopkins university COVID-19 and vaccine data , one world data for population

1) "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_cov                d_19_data/csse_covid_19_time_series/"
  -"time_series_covid19_confirmed_US.csv"
  -"time_series_covid19_confirmed_global.csv"
  -"time_series_covid19_deaths_US.csv"
  -"time_series_covid19_deaths_global.csv
  
2) <https://github.com/owid/covid-19-data/tree/master/public/data/vaccination   s>

3) https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covi   d_19_data/UID_ISO_FIPS_LookUp_Table.csv



### Import COVID019 data

```{r get_data, message=FALSE}

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv", 
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])

```


### Import Vaccine Data


```{r Vaccine_data,message=FALSE}
url_in <- 'https://covid.ourworldindata.org/data/vaccinations/us_state_vaccinations.csv'

US_vaccinations <- read_csv(url_in)
```


### Import world population Data


```{r global_population,message=FALSE}

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

```


## Tidy and Transfrom Data

###Tidy COVID-19 Data

There are 4 data sets: global cases, global deaths, US cases, and US deaths. Each data set has the following columns: Province/State, Country/Region, Lat, Long, and hundreds of date columns. 

Dates are each column which do not look useful. I will pivot dates and take count for each date.

Data Set also contains Lat and Long which are not much useful since I will use country_region to perform analysis.

### Tidy US Cases

```{r tidy_US,message=FALSE}

US_cases <- mutate_at(US_cases, vars(Admin2, Province_State, Country_Region), as.factor) %>%
  rename(County = 'Admin2') %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = 'Date',
               values_to = 'Cases') %>%
  filter(Cases >= 0) %>%
  select(County:Cases) %>%
  mutate(Date = mdy(Date)) %>%
  select(-c(Lat, Long_))


US_deaths <- mutate_at(US_deaths, vars(Admin2, Province_State, Country_Region), as.factor) %>%
  rename(County = 'Admin2') %>%
  pivot_longer(cols = -(UID:Population),
               names_to = 'Date',
               values_to = 'Deaths') %>%
  filter(Deaths >= 0) %>%
  select(County:Deaths) %>%
  mutate(Date = mdy(Date)) %>%
  select(-c(Lat, Long_))

US <- US_cases %>%
  full_join(US_deaths)

# Summarizing cases and deaths by state

US_by_state <- US %>%
  group_by(Province_State, Country_Region, Date) %>%
  summarize(Cases = sum(Cases), Deaths = sum(Deaths), Population = sum(Population)) %>%
  select(Province_State, Country_Region, Date, Cases, Deaths, Population) %>%
  ungroup()

# Summarizing cases and deaths per 100 by state 
# This transformation is done to match with vaccine data which are also in per_hundred

US_by_state_deaths_per_hundred <- US_by_state %>%
  group_by(Province_State, Population) %>%
  summarize(Deaths = max(Deaths)) %>%
  mutate(Deaths_per_hundred = (Deaths/Population)*100) %>%
  arrange(desc(Deaths_per_hundred)) %>%
  filter(Population > 0) %>%
  select(Province_State, Population, Deaths, Deaths_per_hundred) %>%
  ungroup()

```
  
### Tidy Global COVID-19 Data

```{r tidy_global,error=FALSE}

global_cases <- global_cases %>% 
  pivot_longer(cols = -c('Province/State', 
                         'Country/Region', Lat, Long),
               names_to = "date", 
               values_to = "cases") %>%
  select(-c(Lat, Long))


global_deaths <- global_deaths %>% 
  pivot_longer(cols = -c('Province/State', 
                         'Country/Region', Lat, Long),
               names_to = "date", 
               values_to = "deaths") %>%
  select(-c(Lat, Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region', 
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

# Filter out and keep only where the cases are positive
global <- global %>% filter(cases > 0)

global <- global %>%
  unite("Combined_Key", 
        c(Province_State, Country_Region),
        sep = ", ", 
        na.rm = TRUE,
        remove = FALSE)
# Join global  with global population lookup table df and remove unneeded columns

global <- global %>%
	left_join(uid, by = c('Province_State', 'Country_Region')) %>%
	select(-c(UID, FIPS)) %>%
	select(Province_State, Country_Region, date,
		cases , deaths, Population, Combined_Key)

global_by_country <- global %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

global_totals <- global_by_country %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
global_totals
                    
```


### Tidy Vaccine data


```{r ltidy_vaccine_data,error=FALSE,message=FALSE}

US_vaccinations <- US_vaccinations %>%
  select(-c(total_distributed, people_vaccinated, people_fully_vaccinated,
          daily_vaccinations_raw, daily_vaccinations,daily_vaccinations_per_million,
          share_doses_used, total_boosters)) %>%
  rename(Province_State = 'location')

US_by_state_vaccinations_per_hundred <- US_vaccinations %>%
  group_by(Province_State) %>%
  mutate(Province_State = fct_recode(Province_State,
                                     "New York" = "New York State")) %>%
  summarize(people_fully_vaccinated_per_hundred = max(people_fully_vaccinated_per_hundred, na.rm = T),
            total_vaccinations_per_hundred = max(total_vaccinations_per_hundred, na.rm = T), 
            people_vaccinated_per_hundred = max(people_vaccinated_per_hundred, na.rm = T),
            distributed_per_hundred = max(distributed_per_hundred, na.rm = T),
            total_boosters_per_hundred = max(total_boosters_per_hundred, na.rm = T))
  
US_by_state_deaths_vaccinations_per_hundred <- US_by_state_deaths_per_hundred %>%
  full_join(US_by_state_vaccinations_per_hundred) %>%
  filter(Population > 0)

```



## Data Visualization 


### Visualize and Analysis of Global Data


```{r global_trend, results='hide'}

global_by_country %>%
  ggplot(aes(x = date , y = cases)) +
  
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  
  scale_y_log10() +
  
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  
  labs(title = "COVID19 in World", y = NULL)

```


***Observation and Discussion***

The plot shows the trend of cumulative COVID-19 and related fatalities for all over the world. Cases were recorded from early 2020 and it went sharply up and then stabilize after early part of 2022.


Since cumulative plot does not  give the trend; I had plotted the trend graph which displays monthly COVID-19 new cases. We can see new cases keep going up and peak at around Nov/Dec 2021 and and then started falling down.


``` {r global_monthly_trend}
    
trend<- global_totals %>%
  filter(cases > 0) %>%
  mutate(new_cases = cases  - lag(cases , default=0),
         new_deaths = deaths - lag(deaths),
         year_month = format_ISO8601(date, precision = "ym"))  %>%
  group_by(year_month)  %>%
  summarize(new_cases = sum(new_cases), new_deaths = sum(new_deaths)) %>%
  ungroup() 
trend %>%
  ggplot(aes(x = year_month, y = new_cases)) +
  geom_line(aes(color = "new_cases", group = 1)) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths", group = 1)) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "New COVID19 cases/deaths in the world", y = NULL)
```


### Visualize and Analysis US Data

We can see new cases keep going up and peak at around Nov/Dec 2021 and and then started falling down.

```{r visualizing1}

US_by_state %>%
  ggplot(aes(x = Date , y = Cases)) +
  
  geom_line(aes(color = "Cases")) +
  geom_point(aes(color = "Cases")) +
  
  geom_line(aes(y = Deaths, color = "Deaths")) +
  geom_point(aes(y = Deaths, color = "Deaths")) +
  
  scale_y_log10() +
  
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  
  labs(title = "COVID19 in US", y = NULL)
```

***Observation and Discussion***

The US cumulative plot follows similar progression as Global COVID-19 trend 

```{r visualizing2}
trend_US<- US %>%
  filter(Cases > 0) %>%
  mutate(new_cases = Cases  - lag(Cases , default=0),
         new_deaths = Deaths - lag(Deaths),
         year_month = format_ISO8601(Date, precision = "ym"))  %>%
  group_by(year_month)  %>%
  summarize(new_cases = sum(new_cases), new_deaths = sum(new_deaths)) %>%
  ungroup() 
trend %>%
  ggplot(aes(x = year_month, y = new_cases)) +
  geom_line(aes(color = "new_cases", group = 1)) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths", group = 1)) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "New COVID19 cases/deaths in the US", y = NULL)

```

***Observation and Discussion***

New COVID-19 monthly trend also flowing similar as global trend.


### Visualize and Analysise Vaccine Data***



``` {r analyzing}
US_vaccinations %>%
  ggplot( aes(x=date, y=as.numeric(total_vaccinations))) +
  geom_area(fill="Green", alpha=0.5) +
  geom_line(color="Green") +
  ylab("Total immunizations") 
```


***Observations and Discussion***

From the above graph, we can seen vaccination was started early 2021 but it was not available for all for wider public till Mid summer (Jun/July of 2021).


# MODELING

### Modeling of Global data

For this model, a simple linear regression is performed for death for thousand and case per thousand in the global data.


```{r modeling_1}

global_country_totals <- global_by_country %>%
  group_by(Country_Region) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases/population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

mod <- lm(deaths_per_thou ~ cases_per_thou, data = global_country_totals)
summary(mod)
global_country_totals %>% slice_min(cases_per_thou)
global_country_totals %>% slice_max(cases_per_thou)
x_grid <- seq(1, 151)
new_df <- tibble(cases_per_thou = x_grid)
global_country_totals %>% mutate(pred = predict(mod))
global_tot_w_pred <- global_country_totals %>% mutate(pred = predict(mod))
global_tot_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
labs(x = "Cases Per thousand",
       y = "Deaths for thousand",
       title = "Cases Per thousand/ Deaths per thousand  - Linear Model for Globe")
```

***Observations and Discussion***

The Graph clearly shows a correlation between deaths vs cases exihibiting linear pattern.

### Modeling for US Data

For this model, a simple linear regression is performed on the data for COVID-19 vaccination rates and death rates in the US to see if there a significant relationship between the number of people fully vaccinated and number of deaths.The above graph shows the downward trend between fully vaccinated people per 100 members  and deaths per 100 members.


```{r, modeling_2, message=FALSE}

ggplot(data = US_by_state_deaths_vaccinations_per_hundred,
       aes(x = people_fully_vaccinated_per_hundred, 
           y = Deaths_per_hundred,
           label = Province_State)) +
  geom_point(size = .4) +
  geom_text(size = 2, vjust = .5, hjust = -.5) +
  geom_smooth(method = "lm") +
  labs(x = "fully vaccinated per hundred people",
       y = " COVID-19 deaths per hundred people",
       title = "COVID-19 Deaths / Fully Vaccinated People - Linear Model for US")

```


```{r, message=FALSE}
mod1 <- lm(Deaths_per_hundred ~ people_fully_vaccinated_per_hundred, 
          data = US_by_state_deaths_vaccinations_per_hundred)
summary(mod1)

```


***Observations and Discussion***

The summary statistics shows p-value  for people_fully_vaccinated_per_hundred ***2.68e-05** which is way below the bar p-value 0.05 from which we can reject null hypothesis ie there is correlation between Deaths and people fully vaccinated

# CONCLUSION

As we can see from the plots above, COVID cases and deaths over the world and in the US have been both increasing over the time. We reached the highest new cases per day ever in Nov/Dec 2021. Deaths also follow the similar pattern.

From the first liner model on the global data, we can conclude there is linear relationship be between number of cases and number of deaths


From the second model observations, we can safely conclude there is  statistically significant relationship in the US between the number of people in a state that are fully vaccinated for COVID-19 and the number of COVID-19 deaths in that state


# BIAS

There can be a bias in how the data has been collected. For example many people who live in remote area most probably did  not go for the testing. In addition, there are no additional data points other than death and cases.Multiple research show that majority death due to COVID-19 with people with other co-morbidity.


## Session info
```{r, echo=FALSE}
sessionInfo()
```

